---
title: "Effects on geometric distribution when there is growth/decline"
author: "Felix T Petersma"
date: "2023-06-19"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

## Introduction

As the CKMR model currently stands, it contains a contradiction: yearly survival
is the same for all age categories and denoted $\phi$. When a population is 
in constant growth or decline, the age distribution is stable and geometric, 
solely explained by $\phi$. However, even though our simulations all start at
a stable age distribution with growth expectation approximately zero (i.e., 
constant population size), our population dynamics are \emph{stochastic}. This
means that the sex assignments are random with a 50:50 sex ratio; survival is a
Bernoulli trial with survival probability $\phi$; and that, for the complex 
species, fecundity $\beta$ is a random draw from the set \{3, 4, 5, 6\}, i.e., 
$\mathbb{E}[\beta] = 4.5$. This randomness resulted in the majority of our 1000 
realised population histories to be pretty much the same over the years, but 
some showing decline or growth. To accommodate this, we also model yearly growth
rate parameter $r$. At same time, we model the age distribution $f(a;\phi)$ 
assuming a stable age structure, which, if there is any growth or decline, is
not valid. In statistics we often make assumptions that oversimplify reality
(in fact, this is the whole point of a model). The issue at hand is that we 
have a contradiction \emph{within} the model, i.e., we assume a stable age 
structure but also allow the model to fit growth or decline. 

Ideally, one would find a way to come up with an age distribution that 
accommodates this growth or decline. We believe that this is possible \emph{if}
the observed growth or decline is a consequence of a deterministic process, 
e.g., that reproduction 'outperforms' death. That is not the case here; our 
realised growth or decline follows from random processes, chance so to say. 
Maybe there were just a few years that many females were born, resulting in an
abundance of mothers several years later. Or the Bernoulli trials for survival
happened to let many individuals live. Maybe these events are more accurately 
describe as 'shocks'. Either way, a model tries to capture the deterministic 
processes, not random fluctuations. Therefore we see no way to incorporate
the process that drive growth/decline in the age distribution, without 
completely rethinking the way the distribution is determined and estimated 
(e.g., with great white shark study by Richard Hillary et al. from 2018 in 
Nature). Our assumed distribution $f(a)$ is correct in expectation, and thus
we here explore the consequences of keeping it that way when stochastic 
processes lead to growth or decline. It should be noted that, even though the
randomness led to growth or decline, this yearly decline or growth was always
contained in the interval $\pm 0.3\%$. 

For this, we construct a deterministic Leslie matrix for the simple species.
This species has the following population history traits that are important:
+ Males and females both mature at age 10, and maturity is knife-edge;
+ Mature females produce two pups every year with a 50:50 sex ratio;
+ Females mate every year;
+ Males can father multiple litters, but there is only one father to any litter;
+ Survival is the same for all age categories, irrespective of sex;
+ The maximum age is 19, which means that no shark will every reach the age of 20.

We only model the female side of the population, as females exclusively 
determine total reproduction, given that that at least one mature male exists in
the population. This means in effect that every mature female produces one 
female offspring every year. 


## Construction the Leslie matrix

We now present the construction of the Leslie matrix through embedded \texttt{R} 
code. Subsequently, we will do an eigenanalysis to extract the growth rate 
(the dominant eigenvalue) and the asymptotic age distribution (the dominant
eigenvector). In the section after that, we will experiment with stochastic 
shocks to the system and their consequence to the growth rate and age 
distribution. 

```{r simple_leslie}
## DEFINING THE LESLIE MATRIX
## =============================

survival <- 0.8465        # yearly survival for all age categories
exp_fecun <- 2            # expected fecundity
alpha <- 10               # age of maturity (knife-edge)
max_age <- 19             # the maximum age

## Initiate the matrix with zeros
Leslie_simple <- matrix(0, nrow = max_age + 1, ncol = max_age + 1)

## Define the first fecundity row, and only keep fecundity for mature categories
fecun_row <- rep(exp_fecun / 2, times = max_age + 1)
fecun_row[0:(max_age) < alpha] <- 0

Leslie_simple[1, ] <- fecun_row

## Create matrix with survival on the diagonal and zeros elsewhere, for survival
surv_mat <- matrix(0, nrow = max_age, ncol = max_age)
diag(surv_mat) <- survival

Leslie_simple[2:(max_age + 1), 1:max_age] <- surv_mat
```
We now extract the dominant eigenvalue and eigenvector. Sometimes, these
values and vector can be complex, which complicates things a bit. Luckily, our
dominant ones are real, and we extract them below. We capture the real part 
of a complex number using \texttt{Re()}.

```{r simple_eigen_analysis}
## Extract dominant eigenvalue and eigenvector
orig_dom_value <- Re(eigen(Leslie_simple, symmetric=F)$values[1])    
orig_dom_value    # close to 1, which is the expected growth rate 

orig_dom_vector <- Re(eigen(Leslie_simple, symmetric=F)$vectors[, 1])

## Create a plot to show the dominant distribution as a dashed line
plot(y = orig_dom_vector / sum(orig_dom_vector), # divide by sum to be a pdf
     x = 0:19, type = "l", ylab = "Probability density", xlab = "Age", 
     lty = "dashed", lwd = 1.5) 
                                                   
## Now add the 'true' geometric distribution as a solid line
lines(x = 0:19, y = dgeom(0:19,  1 - survival) / pgeom(19, 1 - survival))
```

The expected asymptotic age distribution is near-identical to the assumed 
distribution. We now evaluate what happens when we start adding noise.

## Fecundity is noisy, but unbiased

We now add some random noise to the fecundity. Instead of fecundity being 2, it 
is now $\mathcal{N}(2, 0.2^2)$, so the standard deviation is 10\% of 
the expectation. This did visually nothing of importance, however, that is not 
unexpected as our stochastic is small and has no pattern. The dominant 
eigenvalue remained virtually unchanged (i.e. no growth or decline). 

```{r}
## Create a new fecundity row
set.seed(707707)
fecun_row <- (rep(exp_fecun, times = max_age + 1) + rnorm(20, 0, 0.2)) / 2
fecun_row[0:(max_age) < alpha] <- 0

Leslie_simple[1, ] <- fecun_row

## Re extract the eigen value and vector, and plot
dom_value <- Re(eigen(Leslie_simple, symmetric=F)$values[1])  
dom_value

dom_vector <- Re(eigen(Leslie_simple, symmetric=F)$vectors[, 1])

## Recreate the old plot
plot(y = orig_dom_vector / sum(orig_dom_vector), # divide by sum to be a pdf
     x = 0:19, type = "l", ylab = "Probability density", xlab = "Age", 
     lty = "dashed", lwd = 1.5) 
lines(x = 0:19, y = dgeom(0:19,  1 - (survival / dom_value)) / 
  pgeom(19, prob = 1 - (survival / dom_value)))
## Add the new stuff
lines(x = 0:19, y = dom_vector / sum(dom_vector), col = "red", lty = "dashed")
```

## More or less female offspring

We now explore the effects of a positive bias in the number of offspring of 20\%. 
As total number of offspring per female is fixed at 2, this 20\% increase could 
be the consequence of more females being produced than males. 

```{r}
fecun_row <- rep(exp_fecun / 2, times = max_age + 1) * 1.2 # 20% increase
fecun_row[0:(max_age) < alpha] <- 0

Leslie_simple[1, ] <- fecun_row

## Re extract the eigen value and vector, and plot
dom_value <- Re(eigen(Leslie_simple, symmetric=F)$values[1])  
dom_value

dom_vector <- Re(eigen(Leslie_simple, symmetric=F)$vectors[, 1])

## Recreate the old plot
plot(y = orig_dom_vector / sum(orig_dom_vector), # divide by sum to be a pdf
     x = 0:19, type = "l", ylab = "Probability density", xlab = "Age", 
     lty = "dashed", lwd = 1.5, 
     main = paste0("Yearly growth rate: ", round(dom_value - 1, 3) * 100, "%")) 
lines(x = 0:19, y = dgeom(0:19,  1 - (survival / dom_value)) / 
  pgeom(19, prob = 1 - (survival / dom_value)))
## Add the new stuff
lines(x = 0:19, y = dom_vector / sum(dom_vector), col = "red", lty = "dashed")

```

and what if more males are being born than females, resulting in a decline in 
female offspring of 20\%?

```{r}
fecun_row <- rep(exp_fecun / 2, times = max_age + 1) - 0.2 # 20% decrease
fecun_row[0:(max_age) < alpha] <- 0

Leslie_simple[1, ] <- fecun_row

## Re extract the eigen value and vector, and plot
dom_value <- Re(eigen(Leslie_simple, symmetric=F)$values[1])  
dom_value

dom_vector <- Re(eigen(Leslie_simple, symmetric=F)$vectors[, 1])

## Recreate the old plot
plot(y = orig_dom_vector / sum(orig_dom_vector), # divide by sum to be a pdf
     x = 0:19, type = "l", ylab = "Probability density", xlab = "Age", 
     lty = "dashed", lwd = 1.5, 
     main = paste0("Yearly growth rate: ", round(dom_value - 1, 3) * 100, "%")) 
lines(x = 0:19, y = dgeom(0:19,  1 - (survival / dom_value)) / 
  pgeom(19, prob = 1 - (survival / dom_value)))

## Add the new stuff
lines(x = 0:19, y = dom_vector / sum(dom_vector), col = "red", lty = "dashed")

```

## Survival is higher than

Now we explore the case where, due to randomness, survival is lower than 
expected. 

```{r}
## Set fecundity to original
fecun_row <- rep(exp_fecun / 2, times = max_age + 1) 
fecun_row[0:(max_age) < alpha] <- 0

Leslie_simple[1, ] <- fecun_row

## Now change the survival part
surv_mat <- matrix(0, nrow = max_age, ncol = max_age)
diag(surv_mat) <- survival * 0.7    # 30% decrease

Leslie_simple[2:(max_age + 1), 1:max_age] <- surv_mat

## Re extract the eigen value and vector, and plot
dom_value <- Re(eigen(Leslie_simple, symmetric=F)$values[1])  
dom_value

dom_vector <- Re(eigen(Leslie_simple, symmetric=F)$vectors[, 1])

## Recreate the old plot
plot(y = orig_dom_vector / sum(orig_dom_vector), # divide by sum to be a pdf
     x = 0:19, type = "l", ylab = "Probability density", xlab = "Age", 
     lty = "dashed", lwd = 1.5, 
     main = paste0("Yearly growth rate: ", round(dom_value - 1, 3) * 100, "%")) 
# lines(x = 0:19, y = dgeom(0:19,  1 - survival * 0.5) / pgeom(19, prob = 1 - survival * 0.5))
lines(x = 0:19, y = dgeom(0:19,  1 - (survival * 0.7 / dom_value)) / 
  pgeom(19, prob = 1 - (survival * 0.7 / dom_value)))
## Add the new stuff in red
lines(x = 0:19, y = dom_vector / sum(dom_vector), col = "red", lty = "dashed")

```

Of course, increasing overall survival does nothing to the distribution. Still, 
it is nice to see the empirical confirmation of this. We could change the 
survival for only some categories, but then the age distribution is just going 
to follow that distribution. Let's add some noise to the distribution with a 
small bias in it.

```{r}
## Set fecundity to original
fecun_row <- rep(exp_fecun / 2, times = max_age + 1) 
fecun_row[0:(max_age) < alpha] <- 0

Leslie_simple[1, ] <- fecun_row

## Now change the survival part
surv_mat <- matrix(0, nrow = max_age, ncol = max_age)
diag(surv_mat) <- rep(survival, 19) + rnorm(19, 0.02, 0.1)  # 2% increase, sd=0.1

Leslie_simple[2:(max_age + 1), 1:max_age] <- surv_mat

## Re extract the eigen value and vector, and plot
dom_value <- Re(eigen(Leslie_simple, symmetric=F)$values[1])  
dom_value

dom_vector <- Re(eigen(Leslie_simple, symmetric=F)$vectors[, 1])

## Recreate the old plot
plot(y = orig_dom_vector / sum(orig_dom_vector), # divide by sum to be a pdf
     x = 0:19, type = "l", ylab = "Probability density", xlab = "Age", 
     lty = "dashed", lwd = 1.5, 
     main = paste0("Yearly growth rate: ", round(dom_value - 1, 3) * 100, "%")) 
lines(x = 0:19, y = dgeom(0:19,  1 - survival) / pgeom(19, prob = 1 - survival))

## Add the new stuff
lines(x = 0:19, y = dom_vector / sum(dom_vector), col = "red", lty = "dashed")
```

This looks a little funky, but still, it is not bad, and already associated with
a pretty high yearly growth rate. I guess a question remains is whether the 
age distribution concerns the true age distribution in the population, or the 
sample age distribution. As sampling is random, these two should be the same,
intuitively, but it is worth a discussion. 

In other words, the assumption, in my view, is valid to match the truth. However,
for a given model, we fix survival and we allow for growth. If it then turns out
that the model finds 15\% growth per year, for example, that can only come from
higher fecundity, either through large litters (for the complex) or through the
sex assignment (for simple and complex). 


```{r}
fecun_row <- rep(exp_fecun / 2, times = max_age + 1) + 0.5 # 50% increase
fecun_row[0:(max_age) < alpha] <- 0

Leslie_simple[1, ] <- fecun_row

## Reset the survival part
surv_mat <- matrix(0, nrow = max_age, ncol = max_age)
diag(surv_mat) <- rep(survival, 19) 

Leslie_simple[2:(max_age + 1), 1:max_age] <- surv_mat

## Re extract the eigen value and vector, and plot
dom_value <- Re(eigen(Leslie_simple, symmetric=F)$values[1])  
dom_value

dom_vector <- Re(eigen(Leslie_simple, symmetric=F)$vectors[, 1])

## Recreate the old plot
plot(y = orig_dom_vector / sum(orig_dom_vector), # divide by sum to be a pdf
     x = 0:19, type = "l", ylab = "Probability density", xlab = "Age", 
     lty = "dashed", lwd = 1.5, 
     main = paste0("Yearly growth rate: ", round(dom_value - 1, 3) * 100, "%")) 
lines(x = 0:19, y = dgeom(0:19,  1 - (survival / dom_value)) / 
  pgeom(19, prob = 1 - (survival / dom_value)))

## Add the new stuff
lines(x = 0:19, y = dom_vector / sum(dom_vector), col = "red", lty = "dashed")

```

I think I now finally and truly understand what is the problem here. Basically, 
the model has a fixed survival parameter $\phi$. Since the distribution on age
$f(a; \phi)$ is parameterised only by $phi$, that means that $f(a)$ is fixed
and given, too. This is fine, even when there is population growth or decline.
But the survival parameter was chosen in a way that ensures, for this system
of sharks with a given reproduction, that it the expected growth is zero. If we
then find growth, and survival is still fixed and known, then this growth can 
\emph{only} follow from an increase in female abundance through reproduction. 
This means that the distribution in ages is no longer determined by the 
parameter $phi$ exclusively, but also in some way by growth rate, which is 
a function of the increased fecundity. 

We build on the previous figures and experiment to get the following. It seems 
that the distribution on ages is still geometric but the growth parameter is now
equal to survival divided by growth rate. 

```{r}
## Recreate the old plot
plot(y = orig_dom_vector / sum(orig_dom_vector), # divide by sum to be a pdf
     x = 0:19, type = "l", ylab = "Probability density", xlab = "Age", 
     lty = "dashed", lwd = 1.5, 
     main = paste0("Yearly growth rate: ", round(dom_value - 1, 3) * 100, "%")) 
lines(x = 0:19, y = dgeom(0:19,  1 - (survival / dom_value)) / 
  pgeom(19, prob = 1 - (survival / dom_value)))

## Add the new stuff
lines(x = 0:19, y = dom_vector / sum(dom_vector), col = "red", lty = "dashed")
```

We now test it again, with a massive decline.

```{r}
fecun_row <- rep(exp_fecun / 2, times = max_age + 1) - 0.7 # 70% increase
fecun_row[0:(max_age) < alpha] <- 0

Leslie_simple[1, ] <- fecun_row

## Reset the survival part
surv_mat <- matrix(0, nrow = max_age, ncol = max_age)
diag(surv_mat) <- rep(survival, 19) 

Leslie_simple[2:(max_age + 1), 1:max_age] <- surv_mat

## Re extract the eigen value and vector, and plot
dom_value <- Re(eigen(Leslie_simple, symmetric=F)$values[1])  
dom_value

dom_vector <- Re(eigen(Leslie_simple, symmetric=F)$vectors[, 1])

## Recreate the old plot
plot(y = orig_dom_vector / sum(orig_dom_vector), # divide by sum to be a pdf
     x = 0:19, type = "l", ylab = "Probability density", xlab = "Age", 
     lty = "dashed", lwd = 1.5, 
     main = paste0("Yearly growth rate: ", round(dom_value - 1, 3) * 100, "%")) 
lines(x = 0:19, y = dgeom(0:19,  1 - (survival / dom_value)) / 
  pgeom(19, prob = 1 - (survival / dom_value)))

## Add the new stuff
lines(x = 0:19, y = dom_vector / sum(dom_vector), col = "red", lty = "dashed")

```

It is the same! This is the way to think: imagine that our fecundity in the 
simulation was not 2 but 3. In order to get a population that had zero growth,
we would have to reduce survival. This would make the geometric age distribution
more bendy away from the diagonal, as it meant that fewer animals survived to
the subsequent year. If my model finds growth, and my survival is fixed, then this 
means that the fecundity is higher than two and thus the age distribution should 
be more bendy away from the diagonal. This means that the shape of the age
distribution is the same as if we had no growth with higher fecundity, and thus 
the parameter drive the age distribution should be smaller than survival. In fact,
it turns out that it is survival divided by growth rate, where a  growth rate of
1 means no growth. I think that is exactly what Mark said, which was that the 
log-number-at-age, which is the log of the geometric age distribution, is not
determined by survival but survival minor growth rate, on the log scale. Which 
seems to be the same as what I am finding here. Amazing. 

## Some thoughts the next morning

Okay, what I think is important to know is that the distribution that I am now
finding, the one that is adjusted for growth, is the \emph{asymptotic} 
distribution, and not the current true distribution. This is because, if the
population is in a stable age distribution and then changes, it will approximate
the new stable age distribution but will never reach it.

Also, I still haven't figured out why exactly it is, for any given $\phi$, that
the asymptotic age distribution is geometric with parameter $\phi / r$. 

<!-- ## A potential solution -->

<!-- So, I had a thought about population size, and the information we know.  -->
<!-- Population growth is defined by the function $N_{t}= N_{t_0} \times r ^ {t - t_0}$. -->
<!-- Here, $r$ is a combination of reproductive rate and mortality rate. If there is  -->
<!-- no reproduction, $N_{t+1} = \bar{\phi} N_t = \phi N_t $, since survival is the  -->
<!-- same for all age categories. If not, there would have been some averaging here.  -->
<!-- If there were no survival, then $N_{t+1} = \bar{\beta}N_t$, where  -->
<!-- total reproductive output equal average fecundity rate $\bar{\beta}}$  -->
<!-- multiplied by $N_t$, which is valid only if the age distribution is stable.  -->
<!-- The combination of these two determines $N_t$ is there is fecundity *and*  -->
<!-- mortality, i.e., $N_{t+1} = (\bar{\beta} + \phi )N_t$. If we generalise to more -->
<!-- than one time step, this becomes $N_{t} = (\bar{\beta} + \phi)^{t- t_0}N_{t_0}$. -->
<!-- So, $r = \bar{\beta} + \phi$ -->
